---
title: Play Like A Girl
author: R package build
date: '2022-04-22'
slug: play-like-a-girl
categories: []
tags: []
---

```{r message=FALSE, warning=FALSE, echo=FALSE }
library(tidyverse)
library(waffle)
library(ggrepel)
library(gt)
```

```{r message=FALSE, warning=FALSE, echo=FALSE }
stats <- read_csv("Project2Data.csv")
```
Nearing the end of an abysmal 2021-2022 athletics season, Huskers fans should be entitled to financial compensation for pain and suffering. 

Unless you live under a rock, it’s no secret that Nebraska football, baseball and men’s basketball have not performed well this year. 

And that’s putting it rather nicely. 

Each week, the Husker faithful showed up decked out in red to witness the men’s teams choke over and over again. 

At this point, I would choose to hide under the rock over watching another loss too. 

If the men are playing, the game’s outcome is almost predictable before it starts. Even with a significant early lead, Husker fans hold their breath until the last whistle since the men’s teams have a special ability to invent new ways to lose.

And while all of Nebraska was harping about the men’s performance or lack thereof, very little was said about how successful the Huskers women’s teams have been this year.

How much better are the women compared to the men, you ask?

```{r message=FALSE, warning=FALSE, echo=FALSE }
women <- c("Wins"=84, "Losses"=29)
men <- c("Wins"=30, "Losses"=55,28)

iron(
 waffle(women, 
        rows = 5, 
        title="Nebraska Athletics' Record", 
        xlab="1 square = 1 win or loss", 
        colors = c("#B8336A", "#2B2D42")
        ) + 
   labs(
     x="Women's Teams\n 1 square = 1 win or loss",
     title="Who Runs the World? Girls.", 
     subtitle="The '21- '22 Huskers women's teams saved Nebraska Athletics one game at a time.") + 
   theme(
    plot.title = element_text(size = 25, face = "bold"),
    axis.title = element_text(size = 10),
    plot.subtitle = element_text(size=12),
    plot.caption = element_text(size = 7),
    axis.title.y = element_blank(),
    plot.title.position = "plot"
  ),
  waffle(
   men, 
   rows = 5, 
   colors = c("#007EA7", "#2B2D42", "white")) + 
   labs(
     x = "Men's Teams",
     caption="Source: NCAA Statistics | Graphic by Katy Cowell"
  )
)
```
Statistically, you’ll get 75% more bang for your buck by attending a volleyball, softball or women’s basketball game while there is only 33% chance of a men’s team win. Combined, the 84 wins by the women are just one shy of the 85 total games played by the men. 

The numbers don’t lie; women are carrying Nebraska Athletics this year. 

```{r message=FALSE, warning=FALSE, echo=FALSE }
stackedbarstats <- stats %>%
  select(Team, TotalWins, TotalLosses)
```

```{r message=FALSE, warning=FALSE, echo=FALSE }
widestats <- stackedbarstats %>%
  pivot_longer(
    cols=("TotalWins" | "TotalLosses"), 
    names_to="Type", 
    values_to="Result")
```
Breaking down the total wins and losses by team, women’s sports account for the vast majority of wins to the point where the men’s teams are almost an inverse of their counterparts. 

At the time this writing in mid-April, the softball team is on a hot streak. The Huskers have 18 consecutive wins and they’ve outscored their opponents by 142 runs. Baseball, on the other hand, has lost three games in a row and allowed the other team to outscore them by ten runs.
```{r message=FALSE, warning=FALSE, echo=FALSE }
ggplot() + 
  geom_bar(data=widestats, aes(x=reorder(Team, Result), weight=Result, fill=Type)) + 
  scale_fill_manual(values =c("#2B2D42","#C1CAD6"), name="Result", labels=c("Loss", "Win")) + 
  coord_flip() + 
  labs(
    x="", 
    y="Season Total Wins and Losses", 
    title="The W in Women Stands for Winning", 
    subtitle="A breakdown of the '21- '22 Huskers Athletics records by team.", 
    caption="Source: NCAA Statistics | Graphic by Katy Cowell"
    ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 25, face = "bold"),
    axis.title = element_text(size = 10), 
    plot.subtitle = element_text(size=11), 
    plot.caption = element_text(size = 7),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    ) 
```

```{r message=FALSE, warning=FALSE, echo=FALSE }
HuskerTeams <- c("Football", "Volleyball", "Softball", "Baseball", "Women's Basketball", "Men's Basketball")

teampoints <- stats %>% 
  filter(Team %in% HuskerTeams) %>%
  select(
    Team,
    TotalWins,
    WinPct
    )
```
Just looking at win percentage alone, it’s clear the Nebraska women’s teams are far outperforming the men. 
```{r message=FALSE, warning=FALSE, echo=FALSE }
vball <- teampoints %>% filter(Team == "Volleyball")
sball <- teampoints %>% filter(Team == "Softball")
wbball <- teampoints %>% filter(Team == "Women's Basketball")

ggplot() + 
  geom_point(data=teampoints, aes(x=TotalWins, y=WinPct), color="grey") +
  geom_point(data=vball, aes(x=TotalWins, y=WinPct), color="#B8336A") +
  geom_point(data=sball, aes(x=TotalWins, y=WinPct), color="#B8336A") +
  geom_point(data=wbball, aes(x=TotalWins, y=WinPct), color="#B8336A") +
  geom_text_repel(
    data=teampoints, 
    aes(x=TotalWins, y=WinPct, label=Team)
  ) + 
  labs(
    x="Total Wins", 
    y="Win Percentage", 
    title="A League Of Their Own", 
    subtitle="Nebraska women's basketball, volleyball and softball teams won far more games than the men.", 
    caption="Source: NCAA Statistics | Graphic by Katy Cowell"
    ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 25, face = "bold"),
    axis.title = element_text(size = 9), 
    plot.subtitle = element_text(size=11.5),
    plot.caption = element_text(size = 7),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    ) 
```
```{r message=FALSE, warning=FALSE, echo=FALSE }
rankings <- stats %>% 
  filter(Team %in% HuskerTeams) %>%
  select(Team, BigTenRanking, NationalRanking) %>%
  arrange(BigTenRanking)
```
The Nebraska ladies gave the Huskers' "Sea of Red" something to cheer for both in the Big Ten and nationally. 

This year the volleyball team made it to the national championship, something a Nebraska men’s team hasn’t done since the infamous 1997 football team. After a anxiety inducing five-set match, Nebraska lost to Wisconsin 3-2. Still they ranked second nationally and also second in the toughest conference for volleyball. 

Continuing the winning trend, women’s basketball made it to the March Madness tournament as the No. 8 seed and ended the season ranked 6th in the Big Ten and 23rd in the nation.
 
Softball is also in the nation’s top 25 teams and comes in second in the Big Ten behind Northwestern.

Nebraska Men’s Athletics = Pain. 

The Huskers were the running joke of college football last season and finished last in the conference. A 3-9 record with nine single-digit losses and a 63 point advantage boosted the team’s national ranking but not much else. 

Coming in dead last in the Big Ten, Nebraska men’s basketball was lifeless the entire season up until they had nothing lose.  

```{r message=FALSE, warning=FALSE, echo=FALSE }
rankings %>% 
  gt() %>% 
  cols_label(
    BigTenRanking = "Big Ten Ranking",
    NationalRanking = "National Ranking"
  ) %>%
  tab_header(
    title = "Winners Win",
    subtitle = "The Huskers womens's teams won in the Big Ten and nationally."
  ) %>% tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left", size = 35),
    locations = cells_title("title")
  ) %>% tab_style(
    style = cell_text(color = "black", align = "left", size = 16),
    locations = cells_title("subtitle")
  ) %>%
  tab_source_note(
    source_note = md("**By:** Katy Cowell  |  **Source:** NCAA Statistics")
  ) %>%
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(2)),
       cell_text(weight = "bold", size=12.5)
     )
   ) %>%
  opt_row_striping() %>% 
  opt_table_lines("none")
```
So, what does this all mean? 

It means that it’s about damn time that the Nebraska women’s athletics get the recognition they deserve for how successful they have been this year. 

Their winning record should come with more fans in seats, TV viewership, NIL deals and general support for the programs. These female athletes put their bodies on the line to bleed Husker red and it’s obvious they are outworking the men. 

The results speak for themselves. 

The code that generated this post and these graphs can be found [here] https://github.com/ktcowell/Notebooks/blob/main/Project2THISONE.Rmd.